{\large Gravitational wave sources in our Galactic backyard:}\\{\normalsize Predictions for BHBH, BHNS and NSNS binaries in LISA

### 2021-11-10 paper/tex/sections/results.tex

\subsection{The LISA sensitivity curve and the population of detectable sources}\label{sec:dcos_on_sc}
\begin{figure*}[p]
    \centering
    \includegraphics[width=\textwidth]{fig2_dcos_on_sc.png}
    \caption{The LISA sensitivity curve is shown together with the density distribution of the characteristic strain and the dominant frequency for all detectable sources in our simulations (top) and separated by type (bottom). Contours show the percentage of the population enclosed. The remaining 2\% of the population is shown as dots with a size that scales with the sampling weight. To guide the interpretation, we show reference lines marking where a circular binary would reside for a given distance (diagonal line) and remaining inspiral time (vertical lines), assuming an average chirp mass $\avg{\mathcal{M}_c}$, orientation and sky location. The curved reference line shows the signal expected for an eccentric binary at 30 kpc. The coloured lines in the bottom panels show a contour that encloses 90\% of the population that is circular. LISA verification binaries are overplotted in the top panel (star symbols).  See also Fig.~\ref{fig:dcos_on_sc_ecc_col} and Sec.~\ref{sec:dcos_on_sc} for a discussion. \href{https://github.com/TomWagg/detecting-DCOs-in-LISA/blob/main/paper/figures/fig2_dcos_on_sc.png}{\faFileImage} \href{https://github.com/TomWagg/detecting-DCOs-in-LISA/blob/main/paper/figure_notebooks/sensitivity_curve.ipynb}{\faBook}.}
    \label{fig:dcos_on_sc}
\end{figure*}
We show the expected LISA sensitivity curve \citep{Robson+2019} in Fig.~\ref{fig:dcos_on_sc} and overplot our predictions for the distribution of detectable sources. Eccentric systems emit gravitational waves in multiple harmonic frequencies ($n f_{\rm orb}$, with $n = 2, 3, \dots $). We choose to plot them at the $x$-coordinate that corresponds to the frequency of the harmonic that dominates the SNR. For circular systems, the $x$-coordinate simply corresponds to $2 f_{\rm orb}$. The $y$-coordinate indicates the strength of the signal (or to be more precise, the amplitude spectral density, ASD), including the contribution from \textit{all} harmonics.

For reference, we show dotted lines to indicate where a hypothetical binary system would reside assuming a given distance (diagonal lines) and a fixed remaining inspiral time (vertical lines). For each line we assume the binary is circular and has a chirp mass equal to the average of the sample ($\avg{\mathcal{M}_c}$, annotated in each panel), unless indicated otherwise. We also overplot the LISA verification binaries (star symbols, \citealt{Kupfer+2018}).

@@

We further note that the detectable population is concentrated on the left side of LISA's sensitivity window. The peak is located at a frequency of about $0.4 \unit{mHz}$, which is much lower than the frequency where LISA will be most sensitive (about $10 \unit{mHz}$). This can be understood as an evolutionary effect. DCOs typically form with wide orbits (low frequencies) that would not be detectable yet. Their orbits shrink as they lose angular momentum in the form of gravitational waves leading to an increase of their orbital frequency until they become detectable. Even though the signal emitted by a more compact binary is stronger, such systems are increasingly rare because they evolve faster and faster towards higher frequency as the inspiral accelerates \citep{Peters+1964}. The vertical grid lines show these rapidly decreasing inspiral times at increasing frequencies. Most of the population is thus expected to reside at low frequencies.

In the lower three panels, we show the contributions of the different types of sources. Comparing them, one can observe the shift in the frequency at which the peak is located, at $0.3 \unit{mHz}$, $0.7 \unit{mHz}$ and $1 \unit{mHz}$ for BHBH, BHNS and NSNS systems respectively. This is a result of the difference in chirp mass. A higher mass system can emit at lower frequency and still produce the minimum SNR needed for detection. We note that this effect can be used to distinguish the heavier DCOs that we discuss in this work from WDWD systems, at least probabilistically \citep[e.g.][]{Sesana+2020}. In the same way, this also explains the offset in frequency between the population of sources we predict and the LISA verification binaries (star symbols.

@@

\paragraph{Orbital Frequency} The orbital frequency distributions for BHBHs, BHNSs and NSNSs (Fig.~\ref{fig:fiducial_pdf_distributions}a) peak at progressively increasing frequencies. As already mentioned above, this is because a higher mass DCO at the same distance and eccentricity requires a lower frequency to produce the same signal-to-noise ratio and thus be detected. The distributions appear nearly symmetric, but closer inspections shows that the left hand side is more populated, which can be seen most clearly in the curve for the BHBHs. This is because of the contribution of highly eccentric binaries, which are most abundant in the BHBH population. These systems are still detectable by LISA, despite their low orbital frequency, as the high eccentricity means that the majority of the GW signal is emitted at higher harmonics, where LISA is more sensitive.

@@

A further reason for the preference for low masses is that the population of BHs detectable by LISA primarily come from progenitors that formed from high metallicity gas according to our simulations. Stellar winds are stronger at high metallicity leading to increased mass loss. This affects the mass of the most massive black holes that can be formed \citep{Belczynski+2010}.

@@

In contrast, only $\BHBHqAbovePointEight{}$ of detectable BHBHs are formed with $q > 0.8$ and the distribution peaks around $q = 0.4$. After closer inspection of the binary simulations, we find that the strong stellar winds in our (typically high-metallicity) progenitors are the reason behind this.

@@

At the same time, stellar wind mass loss disfavours the formation of black holes with comparable masses. Such systems would have originated from progenitors that also started with comparable masses. The rather massive secondaries in these systems (especially after they accreted from the primary) experience very strong stellar wind mass loss. This limits the amount by which they can expand. At the same time, wind mass loss (and also stable and more conservative mas transfer in these systems) lead to widening of the orbit. Both effects, the reduced expansion and increased widening of the orbit, tend to prevent the secondaries from being able to fill their Roche lobe. This thus limits the number of systems that experience the reverse mass transfer or common envelope phase needed shrink the orbits and make them tight enough to be detected as gravitational wave sources.

We find that detectable BHNSs have even more unequal mass ratios. Moreover, the mass ratio distribution is bimodal, where the two peaks arise from two distinct formation scenarios. Around two thirds of detectable BHNSs experience at least one common-envelope event, whilst the last third are formed through only stable mass transfer. The first peak at $q = 0.18$ is from systems that experience at least one common-envelope and occurs at the expected mass ratio, which approximately follows the mean BH mass ($\sim 6.5 \unit{M_\odot}$) and NS mass ($\sim 1.2 \unit{M_\odot}$). Yet we also see a second peak at higher mass ratios around $q = 0.34$, which arises from the fraction of the population that underwent only stable mass transfer phases. The stability of mass transfer depends on the mass ratio, leading to a bias for more equal mass ratio systems, resulting in a peak at higher $q$.

\paragraph{Eccentricity} In Fig.~\ref{fig:fiducial_pdf_distributions}d we show the eccentricity distributions. We find that most systems (73\%) will have eccentricities larger than $0.01$ during the LISA mission, which should in principle be detectable according to \citet{Nishizawa+2016}. This means that we will potentially be able to use eccentricity to distinguish these sources from WD sources (see Sec.~\ref{sec:WDWD_distinguish}). We note that several previous studies assumed all systems to be circular when calculating the detection rates \citep[e.g.][]{Liu+2014, Lamberts+2018, Sesana+2020}. See Sec.~\ref{sec:compare_studies} for a discussion of the impact of this assumption.

@@

% \newcolumntype{Y}{>{\centering\arraybackslash}X}
% \begin{table}[tb]
%     \centering
%     \begin{tabularx}{\columnwidth}{l|Y Y Y}
%         \hline
%         \multirow{2}{*}{Threshold} & \multicolumn{3}{c}{Individually detectable harmonics} \\ \cline{2-4}
%         & 0 & 1 & 2+ \\
%         \hline
%         $\mathrm{SNR} > 7$ & 18\%  & 67\% & 15\% \\
%         $\mathrm{SNR} > 3$ &  8\%  & 54\% & 38\% \\
%         $\mathrm{SNR} > 1$ &  4\%  & 30\% & 66\% \\
%     \end{tabularx}
%     \caption{Percentage of systems with individual harmonics above different SNR thresholds.}
%     \label{tab:test}
% \end{table}

@@

%\subsection{Progenitors and formation channels}\label{sec:progenitors_and_formation}
%Our simulations also provide information about the properties of the progenitor systems.  In Figure~\ref{fig:progenitor_properties}, we show the distribution of initial primary masses and initial mass ratios of the progenitors of the LISA detectable DCOs in the top and bottom panel respectively.

%We see that the vast majority of NSNSs are formed with progenitor masses below $20 \unit{M_{\rm \odot}}$ and none are formed with masses greater than $30 \unit{M_{\rm \odot}}$. In contrast, the primary stars in systems that form  BHNSs and BHBHs have very primary progenitor masses that peak around 50\Msun and 80\Msun. We note that the evolution of such massive progenitor stars is poorly understood which is reason for substantial uncertainties in our final predictions which are unfortunately still hard to quantify at present.

%Concerning, the mass ratios we see that, in our fiducial simulations, NSNS tend to form from systems with initial mass ratios close to unity. BHBH and BHNS systems tend to form from systems with initial mass ratios of around 0.3, similar to the trend shown in the final mass final for the DCOs (c.f.\ Fig.~\ref{fig:fiducial_pdf_distributions}c).

%\begin{figure}[htb]
%    \centering
%    \includegraphics[width=\columnwidth]{fig5_progenitor_distributions.pdf}
%    \caption{As Fig.~\ref{fig:fiducial_pdf_distributions}, but for the initial primary mass (top) and initial mass ratio (bottom). Note that the y-axis is log-scaled for the initial primary mass. \href{https://github.com/TomWagg/detecting-DCOs-in-LISA/blob/main/paper/figures/fig5_progenitor_distributions.pdf}{\faFileImage} \href{https://github.com/TomWagg/detecting-DCOs-in-LISA/blob/main/paper/figure_notebooks/fiducial.ipynb}{\faBook}.}
 %   \label{fig:progenitor_properties}
%\end{figure}

\subsection{Formation channels}\label{sec:progenitors_and_formation}
We show the fraction of detectable DCOs that formed through different formation channels in Fig.~\ref{fig:formation_channels}, where the first column corresponds to the fiducial model that we describe in this section. Approximately two thirds of detectable BHBHs are formed through the `only stable mass transfer' channel \citep{vandenHeuvel+2017}, whilst the remaining third are primarly formed through the `classic' channel \citep{Bhattacharya+1991}. Detectable BHNSs follow an inverse pattern, such that around two thirds are formed through the `classic' channel and the rest are mainly formed through only stable mass transfer.

@@

>>>>

\subsection{The LISA sensitivity curve and the population of detectable sources}\label{sec:dcos_on_sc}
\begin{figure*}[p]
    \centering
    \includegraphics[width=\textwidth]{fig2_dcos_on_sc.png}
    \caption{The LISA sensitivity curve is shown together with the density distribution of the characteristic strain and the dominant frequency for all detectable sources in our simulations (top) and separated by type (bottom). Contours show the percentage of the population enclosed. The remaining 2\% of the population is shown as dots with a size that scales with the sampling weight. To guide the interpretation, we show reference lines marking where a circular binary would reside for a given distance (diagonal line) and remaining inspiral time (vertical lines), assuming an average chirp mass $\avg{\mathcal{M}_c}$, orientation and sky location. The coloured lines in the bottom panels show a contour that encloses 90\% of the population that is circular. LISA verification binaries are overplotted in the top panel (star symbols).  See also Fig.~\ref{fig:dcos_on_sc_ecc_col} and Sec.~\ref{sec:dcos_on_sc} for a discussion. \href{https://github.com/TomWagg/detecting-DCOs-in-LISA/blob/main/paper/figures/fig2_dcos_on_sc.png}{\faFileImage} \href{https://github.com/TomWagg/detecting-DCOs-in-LISA/blob/main/paper/figure_notebooks/sensitivity_curve.ipynb}{\faBook}.}
    \label{fig:dcos_on_sc}
\end{figure*}
We show the expected LISA sensitivity curve \citep{Robson+2019} in Fig.~\ref{fig:dcos_on_sc} and overplot our predictions for the distribution of detectable sources. Eccentric systems emit gravitational waves in multiple harmonic frequencies ($n f_{\rm orb}$, with $n = 2, 3, \dots $). We choose to plot them at the $x$-coordinate that corresponds to the frequency of the harmonic that dominates the SNR. For circular systems, the $x$-coordinate simply corresponds to $2 f_{\rm orb}$. The $y$-coordinate indicates the strength of the signal (or to be more precise, the amplitude spectral density, ASD), including the contribution from \textit{all} harmonics.

For reference, we show dotted lines to indicate where a hypothetical binary system would reside assuming a given distance from earth (diagonal lines) and a fixed remaining inspiral time (vertical lines). For each line we assume the binary is circular and has a chirp mass equal to the average of the sample ($\avg{\mathcal{M}_c}$, annotated in each panel), unless indicated otherwise. We also overplot the LISA verification binaries (star symbols, \citealt{Kupfer+2018}).

@@

We further note that the detectable population is concentrated on the left side of LISA's sensitivity window. The peak is located at a frequency of about $0.4 \unit{mHz}$, which is much lower than the frequency where LISA will be most sensitive (about $10 \unit{mHz}$). This can be understood as an evolutionary effect. DCOs typically form with wide orbits (low frequencies) that would not be detectable yet. Their orbits shrink as they lose angular momentum in the form of gravitational waves leading to an increase of their orbital frequency until they become detectable. These systems are increasingly rare because they evolve faster and faster towards higher frequency as the inspiral accelerates, even though the signal emitted by a more compact binary is stronger \citep{Peters+1964}. The vertical grid lines show these rapidly decreasing inspiral times at increasing frequencies. Most of the population is thus expected to reside at low frequencies.

In the lower three panels, we show the contributions of the different types of sources. Comparing them, one can observe the shift in the frequency at which the peak is located, at $0.3 \unit{mHz}$, $0.7 \unit{mHz}$ and $1 \unit{mHz}$ for BHBH, BHNS and NSNS systems respectively. This is a result of the difference in chirp mass. A higher mass system can emit at lower frequency and still produce the minimum SNR needed for detection. We note that this effect can be used to distinguish the heavier DCOs that we discuss in this work from WDWD systems, at least probabilistically \citep[e.g.][]{Sesana+2020}. In the same way, this also explains the offset in frequency between the population of sources we predict and the LISA verification binaries.

@@

\paragraph{Orbital Frequency} The orbital frequency distributions for BHBHs, BHNSs and NSNSs (Fig.~\ref{fig:fiducial_pdf_distributions}a) peak at progressively increasing frequencies. As already mentioned above, this is because a higher mass DCO at the same distance and eccentricity requires a lower frequency to produce the same signal-to-noise ratio and thus be detected. The distributions appear nearly symmetric, but closer inspections shows that the left hand side is more populated, which can be seen most clearly in the curve for the BHBHs. This is due to the contribution of highly eccentric binaries, which are most abundant in the BHBH population. These systems are still detectable by LISA, despite their low orbital frequency, as the high eccentricity means that the majority of the GW signal is emitted at higher harmonics, where LISA is more sensitive.

@@

A further reason that we predict that LISA will not detect high mass systems is that the population of BHs that are detectable by LISA primarily come from progenitors that formed from high metallicity gas according to our simulations. Stellar winds are stronger at high metallicity leading to increased mass loss. This affects the mass of the most massive black holes that can be formed \citep{Belczynski+2010}.

@@

In contrast, only $\BHBHqAbovePointEight{}$ of detectable BHBHs are formed with $q > 0.8$ and the distribution peaks around $q = 0.4$. We find that the strong stellar winds in our (typically high-metallicity) progenitors are the reason behind this.

@@

At the same time, stellar wind mass loss disfavours the formation of black holes with comparable masses. Such systems would have originated from progenitors that also started with comparable masses. The rather massive secondaries in these systems (especially after they accreted from the primary) experience very strong stellar wind mass loss due to LBV-like eruptions. This limits the amount by which they can expand \citep[e.g.][]{vanSon+2021}. At the same time, wind mass loss (and also stable and more conservative mas transfer in these systems) lead to widening of the orbit. Both effects, the reduced expansion and increased widening of the orbit, tend to prevent the secondaries from being able to fill their Roche lobe. This thus limits the number of systems that experience the reverse mass transfer or common envelope phase needed shrink the orbits and make them tight enough to be detected as gravitational wave sources.

We find that detectable BHNSs have even more unequal mass ratios. Moreover, the mass ratio distribution is bimodal, where the two peaks arise from two distinct formation scenarios. Around two thirds of detectable BHNSs experience at least one common-envelope event, whilst the last third are formed through only stable mass transfer. The first peak at $q = 0.18$ is from systems that experience at least one common-envelope and occurs at the expected mass ratio, which approximately follows the mean BH mass ($\sim 6.5 \unit{M_\odot}$) and NS mass ($\sim 1.2 \unit{M_\odot}$). Yet we also see a second peak at higher mass ratios around $q = 0.34$, which arises from the fraction of the population that underwent only stable mass transfer phases. The stability of mass transfer depends on the mass ratio, preferentially forming systems with more equal masses, i.e.\ at higher $q$.

\paragraph{Eccentricity} In Fig.~\ref{fig:fiducial_pdf_distributions}d we show the eccentricity distributions. We find that most systems (73\%) will have eccentricities larger than $0.01$ during the LISA mission, which should in principle be detectable according to \citet{Nishizawa+2016}. This means that we will potentially be able to use eccentricity to distinguish these sources from WD sources, which are expected to have little to no eccentricity (see Sec.~\ref{sec:WDWD_distinguish}). We note that several previous studies assumed all systems to be circular when calculating the detection rates \citep[e.g.][]{Liu+2014, Lamberts+2018, Sesana+2020}. See Sec.~\ref{sec:compare_studies} for a discussion of the impact of this assumption.

@@

@@

\subsection{Formation channels}\label{sec:progenitors_and_formation}
In our fiducial model, approximately two thirds of detectable BHBHs are formed through the `only stable mass transfer' channel \citep{vandenHeuvel+2017}, whilst the remaining third are primarly formed through the `classic' channel \citep{Bhattacharya+1991}. Detectable BHNSs follow an inverse pattern, such that around two thirds are formed through the `classic' channel and the rest are mainly formed through only stable mass transfer.

@@

The fraction of detectable DCOs that formed through different formation channels in other model variations are shown Fig.~\ref{fig:formation_channels}, where the first column corresponds to the fiducial model described above.
